### 2018-05-16 submission/manuscript.Rmd
update to downplay specific bias call outs in discussion

In this study we show that the number of OTUs, error rate, and chimera prevalence depends on polymerase [Figure 1-2 & 4-5]. These differences are important because many diversity metrics rely on the number of OTUs or other measures dependent on error rate and chimera prevalence as part of their metric calculations (e.g. richness). Additionally, the earlier detection of differences in total number of OTUs between polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples. We observed that undetected chimeras that were not identified and removed using standard bioinformatic approaches cause many of these differences. This suggests that specific diversity differences between studies can be attributed to differences in polyermase used. Based on our observations metrics that measure within sample diversity depend on polymerase but this may not be the case for metrics that assess between sample diversity.

@@

The majority of polymerases we studied add small increases in the number of OTUs and chimera prevalence that can be masked by biological differences. The sequence error introduced by the polymerase is small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014]. However, the chimarea prevalence for some polymerases (e.g. Platinum) are relatively large and might be greater than the oberseved biological variation within a specific study. The choice of polymerase can be as important a consideration as either preservation or DNA extraction method used because similar to using different preservation methods or different DNA extraction kits, the type of polymerase can add bias to the observed bacterial community. Although avoiding polymerases that yield a high number of OTUs and chimera prevalence might provide better standardization of results it does come at a cost.

>>>>

In this study we show that the number of OTUs, error rate, and chimera prevalence depends on polymerase used [Figure 1-2 & 4-5]. These differences are important because many diversity metrics rely on the number of OTUs or other measures dependent on error rate and chimera prevalence as part of their metric calculations (e.g. richness). Additionally, the earlier detection of differences in total number of OTUs between polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples. We observed that undetected chimeras that were not identified and removed using standard bioinformatic approaches cause many of these differences. This suggests that specific diversity differences between studies can be attributed to differences in polyermase used. Based on our observations metrics that measure within sample diversity depend on polymerase but this may not be the case for metrics that assess between sample diversity.

@@

The majority of polymerases we studied add small increases in the number of OTUs and chimera prevalence that can be masked by biological differences. The sequence error introduced by the polymerase is also small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014]. The choice of polymerase should be an important consideration in the creation of a 16S rRNA gene sequencing workflow because using different polymerases can add error and bias to the downstream observations. Although standardization of the workflow may partially solve this problem by introducing a consistent bias to all samples, it does come at a cost.

### 2018-05-16 submission/manuscript.tex
update to downplay specific bias call outs in discussion

In this study we show that the number of OTUs, error rate, and chimera
prevalence depends on polymerase {[}Figure 1-2 \& 4-5{]}. These
differences are important because many diversity metrics rely on the
number of OTUs or other measures dependent on error rate and chimera
prevalence as part of their metric calculations (e.g.~richness).
Additionally, the earlier detection of differences in total number of
OTUs between polymerases in the mock versus fecal samples might indicate
that high biomass samples may underestimate the biases present within
low biomass samples. We observed that undetected chimeras that were not
identified and removed using standard bioinformatic approaches cause
many of these differences. This suggests that specific diversity
differences between studies can be attributed to differences in
polyermase used. Based on our observations metrics that measure within
sample diversity depend on polymerase but this may not be the case for
metrics that assess between sample diversity.

@@

The majority of polymerases we studied add small increases in the number
of OTUs and chimera prevalence that can be masked by biological
differences. The sequence error introduced by the polymerase is small
and likely to be smaller than the biological variation within a specific
study, which would be consistent with previous findings for preservation
and DNA extraction methods (Salter et al., 2014; Song et al., 2016; Luo
et al., 2016). However, the chimarea prevalence for some polymerases
(e.g.~Platinum) are relatively large and might be greater than the
oberseved biological variation within a specific study. The choice of
polymerase can be as important a consideration as either preservation or
DNA extraction method used because similar to using different
preservation methods or different DNA extraction kits, the type of
polymerase can add bias to the observed bacterial community. Although
avoiding polymerases that yield a high number of OTUs and chimera
prevalence might provide better standardization of results it does come
at a cost.

>>>>

In this study we show that the number of OTUs, error rate, and chimera
prevalence depends on polymerase used {[}Figure 1-2 \& 4-5{]}. These
differences are important because many diversity metrics rely on the
number of OTUs or other measures dependent on error rate and chimera
prevalence as part of their metric calculations (e.g.~richness).
Additionally, the earlier detection of differences in total number of
OTUs between polymerases in the mock versus fecal samples might indicate
that high biomass samples may underestimate the biases present within
low biomass samples. We observed that undetected chimeras that were not
identified and removed using standard bioinformatic approaches cause
many of these differences. This suggests that specific diversity
differences between studies can be attributed to differences in
polyermase used. Based on our observations metrics that measure within
sample diversity depend on polymerase but this may not be the case for
metrics that assess between sample diversity.

@@

The majority of polymerases we studied add small increases in the number
of OTUs and chimera prevalence that can be masked by biological
differences. The sequence error introduced by the polymerase is also
small and likely to be smaller than the biological variation within a
specific study, which would be consistent with previous findings for
preservation and DNA extraction methods (Salter et al., 2014; Song et
al., 2016; Luo et al., 2016). The choice of polymerase should be an
important consideration in the creation of a 16S rRNA gene sequencing
workflow because using different polymerases can add error and bias to
the downstream observations. Although standardization of the workflow
may partially solve this problem by introducing a consistent bias to all
samples, it does come at a cost.
